Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-16 analysis/paper/paper.qmd
updates of umap figure

# Set point size (highlight idealmodel)
umap_data <- umap_data %>%
  mutate(point_size = if_else(category_lower == "idealmodel", 4.5, 3))

@@

@@

@@

# TODO add the ideal shape graphics to p_umap_by_cluster

ggsave(plot = p_umap_by_category,
       filename = here("analysis/paper/figures/umap_by_category.png"))

@@

library(ggforce)
library(ggrepel)

@@

p_umap_by_cluster <- 
ggplot(umap_data_to_plot, 
       aes(x = x, y = y)) +
  geom_point(
    data = filter(umap_data_to_plot, 
                  is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.8,
    size = 2.5
  ) +
  geom_point(
    data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes"),
    aes(color = cluster),
    size = 5,
    shape = 3,
    show.legend = FALSE
  ) +
  geom_mark_ellipse(aes(color = cluster)) +
  geom_text_repel(data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes",
                  filename == "Sphere"),
    aes(color = cluster,
        label = filename),
    size = 5,
    hjust = 0,
    xlim = c(3, NA),
    ylim = c(2.5, NA),
    nudge_x = 3,
    box.padding = 0.5,
    nudge_y = 1,
    segment.curvature = -0.1,
    segment.ncp = 5,
    segment.angle = 20) +
    geom_text_repel(data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes",
                  filename == "Cube"),
    aes(color = cluster,
        label = filename),
    size = 5,
    hjust = 0,
    xlim = c(3, NA),
    ylim = c(1, NA),
    force = 5,
    nudge_x = 3,
    box.padding = 0.5,
    nudge_y = -5) +
    geom_text_repel(data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes",
                  filename %in% c("Discoid-like", 
                                  "Box",
                                  "Ellipsoid")),
    aes(color = cluster,
        label = filename),
    size = 5,
    nudge_x = 0,
    box.padding = 0.5,
    nudge_y = -2,
    ylim = c(4, 4)) +
  scale_color_manual(
    values = cluster_colors,
    name = "Cluster"
  ) +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded) +  # 保证坐标一致
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )

# TODO add the ideal shape graphics to p_umap_by_cluster

ggsave(plot = p_umap_by_cluster,
       filename = here("analysis/paper/figures/umap_by_cluster.png"))
```

@@

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`). The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying <!-- BM: I think a boxplot would be good to show this to the reader also -->

>>>>

# Set point size (highlight ideal model)
umap_data <- umap_data %>%
  mutate(point_size = if_else(category_lower == "idealmodel", 4.5, 3))

@@

ggplot(umap_data, aes(x = x, y = y)) +
  stat_ellipse(aes(color = cluster), type = "norm", level = 0.95, size = 1) +
  geom_point(aes(color = cluster, size = point_size), alpha = 0.8) +
  geom_text_repel(data = label_data, aes(label = category_lower), size = 4) +
  xlim(x_limits_expanded) +
  ylim(y_limits_expanded) +
  theme_minimal()

@@

library(dplyr)

@@

# 绘图
p_umap_by_category <- p_umap_by_category +
  geom_text_repel(
    data = filter(umap_data, category_lower %in% c("spheroid", "subspheroid")),
    aes(
      x = x,
      y = y,
      label = filename
    ),
    size = 3.5,
    segment.color = "grey",
    box.padding = 0.5,
    point.padding = 0.5
  )

# 显示图形
p_umap_by_category

#ggsave(plot = p_umap_by_category,
       #filename = here("analysis/paper/figures/umap_by_category.png"))

p_umap_by_category

@@

library(ggforce)
library(ggrepel)
library(ggimage)

@@

sphere_data <- filter(umap_data_to_plot, filename == "Sphere")

ellipsoid_data <- filter(umap_data_to_plot, filename == "Ellipsoid")

sphere_data <- sphere_data %>%
  mutate(
    label_x = x + 2.7,    
    label_y = y + 1,      
    mid_x   = x + 1,      
    mid_y   = y + 1      
  )

ellipsoid_data <- ellipsoid_data %>%
  mutate(
    label_x = x + 0.5,    
    label_y = y - 3,      
    mid_x   = x + 0.5,      
    mid_y   = y - 1.6      
  )

p_umap_by_cluster <- ggplot(umap_data_to_plot, aes(x = x, y = y)) +

  geom_point(
    data = filter(umap_data_to_plot, is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.2,
    size = 2.5
  ) +

  geom_point(
    data = filter(umap_data_to_plot, is_idealmodel == "yes"),
    aes(color = cluster),
    size = 4,
    shape = 16,
    alpha = 1,
    shape = 21,                          
    color = "black",                    
    stroke = 1.2,     
    show.legend = FALSE
  ) +

  stat_ellipse(aes(color = cluster), type = "norm", level = 0.95, size = 0.7) +

  geom_point(aes(color = cluster), size = 3) +

  geom_segment(
    data = sphere_data,
    aes(x = x, y = y, xend = mid_x, yend = mid_y, color = cluster)
  ) +
  geom_segment(
    data = sphere_data,
    aes(x = mid_x, y = mid_y, xend = label_x, yend = label_y, color = cluster)
  ) +
  geom_text(
    data = sphere_data,
    aes(x = label_x, y = label_y, label = filename, color = cluster),
    size = 5,
    hjust = 0
  ) +

  geom_segment( 
    data = filter(umap_data_to_plot, is_idealmodel == "yes", 
                  filename == "Cube"), 
    aes(x = x + 2, y = y, xend = x, yend = y, color = cluster) 
  ) + 
  geom_text( 
    data = filter(umap_data_to_plot, is_idealmodel == "yes", 
                  filename == "Cube"), 
    aes(x = x + 2, y = y, label = filename, color = cluster), 
    size = 5, 
    hjust = 0 
  ) +

  geom_text_repel( 
    data = filter(umap_data_to_plot, is_idealmodel == "yes", 
                  filename %in% c("Discoid-like", "Box")), 
    aes(color = cluster, label = filename), 
    size = 5, 
    nudge_x = 0, 
    box.padding = 0.5, 
    nudge_y = -2, 
    ylim = c(4, 4) 
  ) +

    geom_segment(
    data = ellipsoid_data,
    aes(x = x, y = y, xend = mid_x, yend = mid_y, color = cluster)
  ) +
  geom_segment(
    data = ellipsoid_data,
    aes(x = mid_x, y = mid_y, xend = label_x, yend = label_y, color = cluster)
  ) +
  geom_text(
    data = ellipsoid_data,
    aes(x = label_x, y = label_y-0.2, label = filename, color = cluster),
    size = 5,
    hjust = 0.5
  ) +

  scale_color_manual(values = cluster_colors, name = "Cluster") +
  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(x = "UMAP 1", y = "UMAP 2")

image_data <- data.frame(
  x = c(12.5, 5.5, 5.5, 9.5, 15.5),
  y = c(3, 1, 2.8, 3, 3),
  image = c(
    here("analysis", "paper", "figures", "ideal_box.png"),
    here("analysis", "paper", "figures", "ideal_cube.png"),
    here("analysis", "paper", "figures", "ideal_sphere.png"),
    here("analysis", "paper", "figures", "ideal_discoid.png"),
    here("analysis", "paper", "figures", "ideal_ellipsoid.png")
  )
)

p_umap_by_cluster_with_images <- p_umap_by_cluster + 
  geom_image(
    data = image_data,
    aes(x = x, y = y, image = image),
    size = 0.2
  )

ggsave(here("analysis/paper/figures/umap_with_3_images.png"),
       plot = p_umap_by_cluster_with_images,
       width = 10, height = 6, dpi = 300)

@@

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`). The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying
